---
title: "First doc creation - data analysis"
author:
  - name: "Luke Seifert"
    orcid: "0009-0007-4689-8492"
    email: "luse8755@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: today
format: html
editor: visual
---

# Introduction

I am in Blantyre, Malawi, working with Water For People (WFP) - Malawi to collect data on their rural piped water schemes. WFP has built many piped water schemes for rural communities, either from scratch or by upgrading an existing handpump borehole. Once the scheme is handed over to the community, the community is responsible for keeping the system running, encouraging ownership and sustainability, yet some schemes have been struggling to finance their own repairs.

I have been surveying these rural water scheme commities, collecting data on the costs they incur for keeping their water systems running and the monthly revenue they collect via tariffs. By comparing the costs and revenue generated by the schemes, we aim to try and identify the best practices in the rural water sector. The goal is to determine the gaps between costs and revenue, identify any trends between revenue collection and scheme success, and provide stakeholders with the results to encourage data-based decision making. Identifying these trends will help law makers and scheme committies to implement best practices to encourage sustainability of these rural schemes.

My data set will have quantitative data regarding costs of CapEx, CapManEx, and OpEx, quantitative data regarding revenue generated, qualitative data regarding tariff collection methods, and qualitative data regarding scheme sustainability and productivity.

# Methods

## Reading the Data

```{r}

# Load packages
library(tidyverse)
library(dplyr)
library(readr)
library(ggplot2)

# Read data
raw_data <- read_csv(here::here("data/raw/WFP_raw_data_cleaned_7_8.csv"))
```

## Data Exploration Approach

The survey used for data collection was designed by myself in partnership with WFP. Due to resource allocation and a language barrier between myself and the rural communities, we relied on local workers employed by the government to collect the majority of this data. These workers are called "Water Monitoring Assistants", or "WMAs", and they oversee water access infrastructure in these rural areas.

We hosted a training session with the WMAs to show them how to properly deliver this survey to the water scheme committees. Fuel was then provided for the WMAs and they were sent out to their respective areas.

Being several degrees removed from the data we are collecting, the survey responses needed some cleaning up before being able to be analyzed.

## Initial Data Tidying

Before the data set was able to be uploaded into posit.cloud, initial data tidying consisted heavily of fixing obvious data entry errors, addressing inconsistencies or questions left blank, and creating useful data out of response discrepancies.

Some questions were often skipped over and answered partially. Some data recieved was contradictory and needed clarification. Other data was difficult to decipher due to our language barrier. Surveys with these issues were rejected and sent back to the corresponding WMA for re-submission.

The resulting tidied data set has been uploaded to the data/raw folder, and is ready for further data tidying and analysis.

```{r}
#my csv has too many rows present - I want to delete all the empty rows
```

# Results

## Initial Data Exploration

```{r}

#adding data to view first few rows of raw data
#head(raw_data)
dim(raw_data)
#spec(raw_data)

  

```

\[This will be the core of your analysis with specific requirements\]

# Conclusions

## Summary of Findings

## Questions and Next Steps
